---
title: "Measurement Error"
subtitle: POST 8000  -- Foundations of Social Science Research for Public Policy
author: Steven V. Miller
institute: Department of Political Science
titlegraphic: /Dropbox/teaching/clemson-academic.png
date: 
fontsize: 10pt
output:
 beamer_presentation:
    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm-latex-beamer.tex
    latex_engine: xelatex
    dev: cairo_pdf
    fig_caption: true
    slide_level: 3
make149: true
mainfont: "Open Sans"
titlefont: "Titillium Web"
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=TRUE, warning=F)
knitr::opts_chunk$set(fig.path='figs/')
knitr::opts_chunk$set(cache.path='cache/')

knitr::opts_chunk$set(
                  fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      }
                  )
```

```{r loadstuff, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
library(tidyverse)
library(stevemisc)
library(post8000r)

```

# Measurement Error
## Introduction
### Goal for Today

*Discuss what to avoid in your research design (i.e. measurement error).*


### On Error

Error is a fundamental part of a lot of research.

- e.g. random sampling error, random differences between treatment and control.

Thus, KKV argue to reduce error to "manageable proportions." (p. 150)

## Typss of Error
### The Two Types of Error

1. Systematic
    - Definition: a chronic, consistent distortion of a measure in a predictable way.
2. Random
    - Definition: distortions in the recorded value that is neither a function of the “true” signal being measured, nor predictable.

### Common Types of Systematic Error

1. Selection bias
2. Omitted variable bias
3. Related: endogeneity


### Common Types of Random Measurement Error

1. Miscoding
2. Including irrelevant variables (i.e. garbage can modeling)

### Missing Data

Missing data is a unique kind of measurement error. Types:

1. Missing completely at random (MCAR)
2. Missing at random (MAR)
3. Missing not at random (MNAR)

## Effects of Error
### Effects of Systematic Measurement Error

- Invalidity
- Bias in our descriptive inferences
- Potential bias in our causal inferences
- Type 1 error

### Effects of Random Measurement Error


- Inefficiency
- Unreliability
- Type 2 error
- Increased variance

### Effects of Missing Data

MCAR: inefficiency

- Typically requires no correction unless missingness > 10% of the data.

MAR: bias, but fixable

- Missingness depends on known/measured quantities.
- Underlying assumption of multiple imputation.

MNAR: bias, but unsalvagable

- Silver lining: we typically assume MNAR is actually MAR.
- We can eventually find the reasons why

### Some Caveats on Random Measurement Error

- Random error in the DV: creates more noise, greater difficulty in extracting a signal (huge errors)
- Random error in an IV: collapses effect gradually to zero
        - Potential spillover with other IVs, if they're correlated.


## Discussion
### Discussion

1. What's worse: Type 1 or Type 2 error?
2. When is omitted variable bias not a bias at all?
3. How do you "know" when you have a biased measure?
