---
title: "Logistic Regression"
subtitle: POST 8000  -- Foundations of Social Science Research for Public Policy
author: Steven V. Miller
institute: Department of Political Science
titlegraphic: /Dropbox/teaching/clemson-academic.png
date: 
fontsize: 10pt
output:
 beamer_presentation:
    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm-latex-beamer.tex
    latex_engine: xelatex
    dev: cairo_pdf
    fig_caption: true
    slide_level: 3
make149: true
mainfont: "Open Sans"
titlefont: "Titillium Web"
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=TRUE, warning=F)
knitr::opts_chunk$set(fig.path='figs/')
knitr::opts_chunk$set(cache.path='cache/')

knitr::opts_chunk$set(
                  fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      }
                  )
```

```{r loadstuff, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
library(tidyverse)
library(stevemisc)
library(stevedata)
library(knitr)

Timings <- readRDS("run-times.rds")

```

# Logistic Regression
## Introduction
### Goal for Today

*Discuss logistic regression, perhaps the most common form of regression.*

## The Biggest Practical Obstacle to OLS
### OLS

OLS has a ton of nice properties.

- Best linear unbiased estimator (BLUE)
- Simple to execute and interpret.

It'd be a *shame* if something were to happen to one of your assumptions.

### The Problem of Binary DVs

The biggest problem you'll encounter will concern your DV.

- OLS assumes the DV is distributed normally.

You'll most often encounter DVs that are binary.

- Candidate won/lost.
- Citizen voted/did not vote.
- Program succeeded/failed.
- War happened/did not happen.

Most social/political phenomena are typically "there"/"not there."

### The Problem of Binary DVs

Observe this simple data frame, `fakeLogit` from `{stevedata}`.

```{r}
fakeLogit
```

### The Problem of Binary DVs

This simple `fakeLogit` data frame is simulated where:

- `x` is a five-item ordered categorical variable [0:4].
- `y` is a binary variable with only 0s and 1s.
- The effect of a one-unit increase of `x` on `y` is 1.4.
- `y` is estimated to be -2.8 when `x` is 0.

*Importantly*: responses are simulated from a binomial distribution.

- Seed is set for 100% reproducibility.


### The Problem of Binary DVs

Here's what OLS produces.

```{r}
M1 <- lm(y ~ x, fakeLogit)
broom::tidy(M1) %>%
    mutate_if(is.numeric, ~round(., 2)) %>%
    kable(., "markdown")
```


Not even close.

###

```{r fitted-resid-plot-m1, echo=F, eval=T, fig.width = 14, fig.height = 8.5, warning = F, message = F}
fakeLogit %>%
    mutate(fitted = fitted(M1),
           resid = resid(M1)) %>%
    ggplot(.,aes(fitted, resid)) + geom_point()  +
    theme_steve_web() +
    labs(x = "Fitted Values",
         y = "Residuals",
         title = "The Fitted-Residual Plot from the OLS Model We Just Ran",
         subtitle = "No fitted-residual plot from an OLS model should look like this.")

```


### 

```{r qqplot-m1, echo=F, eval=T, fig.width = 14, fig.height = 8.5, warning = F, message = F}
fakeLogit %>%
    mutate(resid = resid(M1)) %>%
    ggplot(.,aes(sample=resid)) + 
    geom_qq()  + theme_steve_web() +
    labs(x = "Theoretical Quantiles",
         y = "Sample Quantiles",
         title = "The Q-Q Plot from the OLS Model We Just Ran",
         subtitle = "The Q-Q plot thinks you messed up too, pay careful attention to the middle of the plot as well.")
```

### The Right Tool for the Right Job

```{r}
M2 <- glm(y ~ x, fakeLogit, family=binomial(link = "logit"))
broom::tidy(M2) %>%
    mutate_if(is.numeric, ~round(., 2)) %>%
    kable(., "markdown")
```

## Unpacking Logistic Regression
### What We Just Did

This was a logistic regression.

- The coefficient tells us the effect of a unit change in `x` on the *natural logged odds of `y`*.

Let's unpack this piece by piece.

### Odds


You typically hear of **odds** in the world of sports betting.

- It's closely linked with probability.

Given some probability *p* of an event occurring, the odds of the event equal:

$$
\textrm{Odds} = \frac{p}{1-p}
$$

Ever hear of something like "the odds are 4 to 1 against" an event occurring?

- Translation: for every five trials, we expect 1 occurrence to 4 non-occurrences, on average.
- Odds >1 = more "successes" than "failures."

### Probability and Odds in Our Data

```{r}
fakeLogit %>%
    group_by(x) %>%
    summarize(sum = sum(y),
              length = length(y),
              p = sum/length,
              # q is often substituted as notation for 1 - p
              q = 1 - p, 
              odds = p/q) -> summaries
```

### Probability and Odds in Our Data

```{r, echo=F}
summaries %>%
    mutate_if(is.numeric, ~round(., 2)) %>%
    kable(.,"markdown")
```

Tell me if you see a pattern beginning to emerge.

### Odds Ratio and Percentage Change in Odds

One way of thinking about change in odds is the odds ratio.

- Simply: the odds of $y$ in one category over the odds from the previous category.

A percentage change in odds is also a useful way of seeing a consistent pattern emerge.

- Simply: the difference in odds for a one-unit increase in `x` over odds of lower category.

### Odds Ratio and Percentage Change in Odds

\small
```{r}
summaries %>%
    mutate(oddsr = odds/lag(odds, 1),
           pcodds = (odds - lag(odds, 1))/lag(odds, 1)*100) -> summaries
```
\normalsize

### Percentage Change in Odds

```{r, echo=F}
summaries %>%
    mutate_if(is.numeric, ~round(., 2)) %>%
    kable(.,"markdown")
```

### Logit (Natural Logged Odds)

Another more sophisticated/flexible way: logits (i.e. natural logged odds).

- These are natural logarithmic transformations (of base $e$) of the odds.

### Logit (Natural Logged Odds)

```{r, echo=T}
summaries %>%
    mutate(logit = log(odds)) %>%
    mutate_if(is.numeric, ~round(., 2)) %>%
    kable(.,"markdown")
```

Now do you see it?


### Compare M2 with the Previous Table

```{r, echo=F}
M2 <- glm(y ~ x, fakeLogit, family=binomial(link = "logit"))
broom::tidy(M2) %>%
    mutate_if(is.numeric, ~round(., 2)) %>%
    kable(., "markdown")
```

## Properties of Logistic Regression Output
### Properties of Logistic Regression

You can always "backtrack" a logistic regression coefficient.

- Exponentiating a logistic regression coefficient returns an odds ratio. Observe:

```{r}
pull(exp(broom::tidy(M2)[2,2]))
```

- You can subtract 1 from the exponentiated coefficient, and multiply it by 100.

```{r}
pull(100*(exp(broom::tidy(M2)[2,2]) - 1))
```

That's the percentage change in odds.

### Properties of Logistic Regression

If you internalize the relationship between probability and odds, you can even return a probability estimate from a logistic regression.


$$
	\textrm{Probability} = \frac{\textrm{Odds}}{1 + \textrm{Odds}} 
$$

In R, for when $x$ = 0, $pr(y = 1 | x = 0)$:.

```{r}
yintercept <- pull(broom::tidy(M2)[1,2])

exp(yintercept)/(1 + exp(yintercept))
```

### Properties of Logistic Regression

For larger/more complex models, *resist the urge to do this by hand. *

- But you could if you knew what you were doing.

For example, here's the probability of $y = 1$ for when $x = 2$:

```{r}
yintercept <- pull(broom::tidy(M2)[1,2])
betax <- pull(broom::tidy(M2)[2,2])

exp(yintercept + 2*betax)/(1 + exp(yintercept + 2*betax))
```

Save this train of thought for when we get to the week on making the most of regression.

### Properties of Logistic Regression

The logistic function is still monotonic, if not exactly linear.

- Interestingly, logistic function is close to linear when $p$ is between .2 and .8.

Think of the logistic regression function as a natural logged odds of "success."

- Recall: dummies are a unique case of a categorical variable.


### Properties of Logistic Regression

*Statistical significance assessments are effectively identical to OLS*.

- Caveat: inference is done via *z*-score and not a *t*-statistic.

tl;dr for why: OLS has both a mean and variance to estimate and the variance is independent of the mean.

- In logistic regression, there's really just one parameter $p$ and not two.
- Basically, the variance with binary data is a function of the mean (i.e. $p(1-p)$).

## Model Fit for Logistic Regression
### Model Fit for Logistic Regression

**Deviance** is the estimate of model fit, not $R^2$.

- Similar to a chi-square analysis.
- i.e. how well does the fitted value ($\hat{y}$) "fit" to the observed value of $y$.

Bigger the difference (or "deviance"), the poorer the fit of the model.

- This will allow you to do some model comparisons with multiple IVs.

### Maximum Likelihood Estimation (MLE)

MLE replaces the OLS principle.

- OLS: draw a line that minimizes the sum of squared  residuals.
- MLE: draw a line that results in the smallest possible deviance.

*This is done iteratively*.

- It's one reason why logistic regression models are slower than linear models.

###

```{r  glm-lm-times-plot, echo=F, eval=T, fig.width = 14, fig.height = 8.5, warning = F, message = F}
Timings %>%
    tbl_df() %>%
    group_by(model) %>%
    ggplot(.,aes(value,  fill=model)) +
    geom_density(color="black") + theme_steve_web() +
    labs(x = "Elapsed Time (in Seconds)",
         y = "Density",
         title = "The Distribution of Run Times for a Linear Regression and Logistic Regression (on the Same Data)",
         fill = "",
         caption = "Data: see R Markdown file for underlying data.",
         subtitle = "GLMs (like logistic regression) take discernibly longer to run, and you'll notice it more in more complicated models.")
```

## Conclusion
### Conclusion

If your DV is binary, use a logistic regression and not OLS.

- Statistical signifiance may not change, but that's also not the point.
- Binary DVs violate the assumptions of OLS and produce misleading estimates.
    - *That's the point*.

The process really doesn't change much.

- Inference is done via standard normal distribution, not Student's t-distribution.
- *Coefficients communicate changes in the natural logged odds of $y$ for a one-unit change in $x$.*

This may take some time, but you'll get used to it. I promise.