---
title: "Random Assignment and Experiments"
subtitle: POST 8000  -- Foundations of Social Science Research for Public Policy
author: Steven V. Miller
institute: Department of Political Science
titlegraphic: /Dropbox/teaching/clemson-academic.png
date: 
fontsize: 10pt
output:
 beamer_presentation:
    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm-latex-beamer.tex
    latex_engine: xelatex
    dev: cairo_pdf
    fig_caption: true
    slide_level: 3
make149: true
mainfont: "Open Sans"
titlefont: "Titillium Web"
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=TRUE, warning=F)
knitr::opts_chunk$set(fig.path='figs/')
knitr::opts_chunk$set(cache.path='cache/')

knitr::opts_chunk$set(
                  fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      }
                  )
```

```{r loadstuff, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
library(tidyverse)
library(stevemisc)
library(post8000r)

```

# Random Assignment and Experiments
## Introduction
### Goal for Today

*Elaborate on the 'gold standard' for causal inference research, with its benefits and limitations.*

## Observational vs. Experimental Research
### Observational vs. Experimental Research

Observational research: involves a comparison of units subjected to different treatments.

- More common, more flexible. But difficult to isolate causal effects.

Experimental research: units under study are randomly assigned to treatments.

- satisfies key questions about observational research design


### Experiments

Experiments are more effective at addressing causality.

- Want to explain social phenomena like medical researchers testing therapeutic care.
- Satisfies insights from Rubin's potential outcomes framework
- Researcher control over conditions isolates confounding systematic factors.
- Random assignment isolates systematic differences from random differences.

### Random Assignment and Causal Inference

Random assignment (e.g. coin-flipping) means each unit has same chance to be in particular group (e.g. control or treatment).

- All groups are equal in expectation, beyond treatment(s).
- Control group behaves as treatment group(s) without treatment, as counterfactual.
- Various tests (e.g. *t*-test) can assess differences between control and treatment(s).


### The Utility of Experiments (Roth, 1995)

1. *"Speaking with theorists":* provides an experimental test for a theoretical model.
2. *"Searching for facts":* generates new data to explore aspects of previous experiments.
3. *"Whispering in the ears of princes":* isolates causal effects, certainly of interest to policymakers.


### Types of Experiments

There are numerous ways of assessing causal effects. One typology:

1. "Between subjects": units randomly assigned to distinct treatment/control groups.
2. "Within subjects": units observed before and after receiving a treatment.

## Validity Concerns
### Validity Concerns

Experiments ideally maximize internal validity, if (possibly) at the expense of external validity.

### Internal Validity

Internal validity: stimulus faithly administered, as implemented in the design. Concerns:

- Noncompliance
- Attrition


### External Validity

External validity: results generalizable from the "lab" to the "real world." Concerns:

- Convenience sampling (esp. college students)
- Hawthorne effect

## Types of Experiments
### Types of Experiments

Experiments are super-flexible. Some types you'll encounter:

1. Lab experiments
    - Maximize internal validity, prioritized over external validity
    - Typically prone to convenience sampling.
2. Survey experiments
    - Balance internal/external validity concerns
    - Typically higher *n* with more representativeness
    - Concerns: spillover, less agency over treatment
3. Field experiment
    - Same pros/cons as survey experiments, but with typically less control over treatment administration.
    - Cons (spillover, treatments) even more pronounced
4. Natural experiment
    - i.e. an exogenous shock to a panel design
5. Quasi-experiment
    - Treatments/controls with no randomization, or control over the treatment.

## Discussion   
### Discussion

1. What's more important: internal or external validity?
2. Think of your research: what type of experiment is feasible for you? What would it look like?

