---
title: "Descriptive Inference"
subtitle: POST 8000  -- Foundations of Social Science Research for Public Policy
author: Steven V. Miller
institute: Department of Political Science
titlegraphic: /Dropbox/teaching/clemson-academic.png
date: 
fontsize: 10pt
output:
 beamer_presentation:
    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm-latex-beamer.tex
    latex_engine: xelatex
    dev: cairo_pdf
    fig_caption: true
    slide_level: 3
make149: true
mainfont: "Open Sans"
titlefont: "Titillium Web"
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=TRUE, warning=F)
knitr::opts_chunk$set(fig.path='figs/')
knitr::opts_chunk$set(cache.path='cache/')

knitr::opts_chunk$set(
                  fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      }
                  )
```

```{r loadstuff, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
library(tidyverse)
library(stevemisc)
library(stevedata)

```

# Descriptive Inference
## Introduction
### Goal for Today

*Introduce students to basic descriptive inference.*

### What Makes Good Scientific Research?

1. Generates inference.
2. The research procedures are transparent/public.
3. The conclusions are "uncertain" (i.e. includes reasonable estimates of uncertainty).
4. The content is the method.

## Interpretation and Inference
### Interpretation and Inference

KKV spend much of chp. 2 contrasting "interpretation" with "inference."

- "Interpretation" seeks accurate summaries of historical detail, placing them in historical context in which we can understand their meaning.

KKV want something more with "inference."

- i.e. to use the facts we know to learn more about the facts we don't yet know.

## A Generalized Model of Research Design
### A Generalized Model of Research Design

KKV build on the "model" metaphor to formalize research design. Definitions first:

- **Data collection**: a wide range of methods for gathering information, ideally connected to the question we want to answer.
- **Variable**: an empirical measurement of a characteristic, typically denoted as $y$.
- **Unit**: a level of analysis (e.g. individual people, countries, counties, etc.)
- **Observation**: an individual observation of *y*, typically denoted like $y_i$.


### An Example from the General Social Survey (2018)

My `{stevedata}` package has a `gss_spending` data frame from the GSS in 2018.

- **Data collection**: a probability sample of individuals living in the U.S.
- **Variable**: (take your pick of various demographic attributes and attitudes toward spending programs)
- **Unit**: individuals living in the U.S.
- **Observation**: a numeric value (again, take your pick, type `?gss_spending` in R for more information).
    - example: `race` = 1 for white people, 2 for black people, 3 for other.


### Summarizing Historical Detail

There are any number of ways of summarizing your variable into a *statistic*, but two principles apply:

1. The statistic should focus on the outcome the researcher wishes to describe or explain.
2. *The goal of the model is to simplify*.


### Descriptive Inference

"Descriptive inference" is the process of understanding the unobserved from the observed. Two governing principles:

1. There are *systematic* differences we can explain by reference to something fundamental/predictable.
2. There are still *nonsystematic* (or *stochastic*) differences from stuff we can't explain.

Our goal is to distinguish the former from the latter.

## Discussion
### Discussion

1. How do we evaluate descriptive inference?
2. What can we do about stochastic/nonsystematic differences in our variable?
3. What is the difference between "realized" and "random" in our variables?
4. What can interpretative methods teach us about inference?