---
title: "Regression Discontinuity Design"
subtitle: POST 8000  -- Foundations of Social Science Research for Public Policy
author: Steven V. Miller
institute: Department of Political Science
titlegraphic: /Dropbox/teaching/clemson-academic.png
date: 
fontsize: 10pt
output:
 beamer_presentation:
    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm-latex-beamer.tex
    latex_engine: xelatex
    dev: cairo_pdf
    fig_caption: true
    slide_level: 3
make149: true
mainfont: "Open Sans"
titlefont: "Titillium Web"
---

```{r setup, include=FALSE, cache=F, message=F, warning=F, results="hide"}
knitr::opts_chunk$set(cache=TRUE, warning=F)
knitr::opts_chunk$set(fig.path='figs/')
knitr::opts_chunk$set(cache.path='cache/')

knitr::opts_chunk$set(
                  fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      }
                  )
```

```{r loadstuff, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
library(tidyverse)
library(stevemisc)
library(post8000r)
library(knitr)

```

# Regression Discontinuity Design
## Introduction
### Goal for Today

*Discuss regression discontinuity design (RDD) as a means to causal inference.*

### Introduction

RDD allows us to estimate causal treatment effects in non-experimental settings.

- It exploits precise knowledge of the rules determining treatment.
- Identification is based on the idea that some rules are arbitrary and provide good quasi-experiments.

Recent flurry of applied economics research using RDD:

- Seemingly mild assumptions (Hahn, Todd, and van der Klaauw 2001)
- More credible than other non-experimental identification strategies (Lee 2008) 

### Origins: Thistlethwaite and Campbell (1960)

Question: what is the effect of merit awards on future academic outcomes?

- However, awards were allocated on the basis of test scores.

There's a cutoff point $c$.

- Below which: no award. Above which: award.

### Origins: Thistlethwaite and Campbell (1960)

Couldn't you just compare those with/without award?

- No: factors that influence the test score are also related to future academic outcomes.

*But* you could compare individuals just above and below the cutoff point $c$.

- That gives you the estimated causal effect of a particular treatment.

###

```{r rdd60-example, echo=F, eval=T, fig.width = 14, fig.height = 8.5, warning = F, message = F}
# RDD60 %>%
#   filter(testscore %in% as.character(1:20)) %>%
#   mutate(testscore = as.integer(testscore)) %>%
#   # select(testscore:intellectualism) %>%
#   select(-nmeritschol, -nresp, -nsampl) %>%
#   gather(variable, value, -meritcerc, -testscore) %>%
#   ggplot(aes(x = testscore, y = value, group = meritcerc)) +
#   facet_wrap(~ variable, ncol = 2, scale = "free_y") +
#   geom_vline(xintercept = 10, colour = "white", size = 2) +
#   geom_point() +
#   geom_smooth(method = "lm") +
#   labs(x = "Aribtrary test scores of students in arbitrary units",
#        y = "")

RDD60 %>%
    filter(testscore %in% as.character(1:20)) %>%
    mutate(testscore = as.integer(testscore))  %>%
    ggplot(.,aes(testscore, schol, group=meritcerc)) +
    theme_steve_web() +
    geom_vline(xintercept = 10.5, color = "gray77", size=2) +
    geom_point() +
    annotate("label", x = 0, y = 65, label = "No Certificate of Merit", hjust = 0, family="Open Sans") +
    annotate("label", x = 20, y = 65, label = "Certificate of Merit Winners", hjust = 1, family="Open Sans") +
    geom_smooth(method = "lm") +
    scale_x_continuous(breaks = seq(1:20)) +
    labs(x = "Aptitude Score Interval (1:20)",
         y = "Percent in Interval Reporting Scholarships",
         title = "An Example of a Regression Discontinuity Design from Thistlethwaite and Campbell (1960)",
         subtitle = "Notice the different effect for both groups, but more importantly notice the discontinuity.",
         caption = "Data: Thistlethwaite and Campbell (1960)")
```

<!-- ### Psst... This is a Simple Interaction -->

<!-- ```{r, results="asis"} -->

<!-- RDD60 %>% -->
<!--     filter(testscore %in% as.character(1:20)) %>% -->
<!--     mutate(testscore = as.integer(testscore)) %>% -->
<!--      lm(schol ~ testscore*meritcerc, data = .) %>% -->
<!--     broom::tidy() %>% -->
<!--     mutate_if(is.numeric, list(~round(.,3))) %>% -->
<!--     kable(.) -->

<!-- ``` -->


<!-- ### Psst... This is a Simple Interaction -->

<!-- ```{r, results="asis"} -->

<!-- RDD60 %>% slice(11:12) %>% -->
<!--     select(1:3, schol) %>% -->
<!--     kable(.) -->

<!-- ``` -->

## Intuition and Identification
### The Intuition Behind RDD (In This Case)

- Assignment mechanism is known.
- The probability of treatment jumps to 1 if test score $\ge c$:
    - We call this a **sharp** discontinuity.
- Individuals cannot manipulate their assignment variable.
- Individuals near cutoff are comparable/similar, save for that one distinction.

Thus, the discontinuous jump in outcome variables at $c$ amounts to the causal effect of merit award.


### Identification in RDD

First:

- All other factors determining the outcome variable should be evolving smoothly with respect to the assignment variable.
- If other factors (variables) also jump at the cutoff point, then the estimates of treatment effect will be biased.

Second:

- Since RD estimate requires data away from $c$, the estimate will be dependent on a chosen functional form.

## Sharp and Fuzzy
### Sharp and Fuzzy RDDs

Thistlethwaite and Campbell (1960) is a clear case of a "sharp" RDD.

- Discontinuity precisely determines treatment (i.e. $p(treatment)$ jumps to 1 at $c$).
- Equivalent to random assignment.

Social Security is a nice example of a sharp design in the wild.

- You can elect to take it earlier, but after a certain threshold, you have to take it.
- (Or at least there's no benefit from delaying payment)

### Sharp and Fuzzy RDDs

"Fuzzy" RDDs are when treatments are just highly correlated with treatments.

- Probability of treatment jumps by less than one when $x$ crosses the threshold $c$.

"Fuzzy" situations make the assignment rule-as-IV an appropriate solution.

## Conclusion
### Conclusion

A regression discontinuity design is a way of undertaking causal inference, usually of some policy intervention.

- It can provide robust, convincing estimates of causal impacts under fairly weak conditions or minimal assumptions.

The nature of the intervention will determine whether an RDD is appropriate.

- Caveat: even when it is, data demands are often great
